---
title: 'FGC_0005: HCT116 - 2 vs 3 replicates'
author: "Alex Kalinka"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    code_folding: hide
    depth: 3
    highlight: tango
    number_sections: true
    theme: spacelab
    toc: true
    toc_float: true
---

# Setup

```{r,echo=F}
options(warn=-1)

suppressMessages(library(knitr))
suppressMessages(library(tidyr))
suppressMessages(library(magrittr))
suppressMessages(library(ggplot2))
suppressMessages(library(dplyr))
suppressMessages(library(fgcQC))
suppressMessages(library(GGally))
suppressMessages(library(locfdr))

filter <- dplyr::filter
select <- dplyr::select
rename <- dplyr::rename
mutate <- dplyr::mutate


# Shared hits.
hits.shared <- read.csv("../data/fgc_0005_depletion_hit_table.csv", stringsAsFactors = F) %>%
  filter(hit_type == "Shared")


# Data kept locally.
mageck_files <- list.files("~/data/az_cruk/cdc7", 
                           pattern = "Treatment_vs_Control.gene_summary.txt", recursive = T, full.names = T)

desir_files <- list.files("~/data/az_cruk/cdc7", 
                          pattern = "DesirabilityGeneRanking.tsv",recursive = T, full.names = T)

read_gene_summ <- function(files, type = "mageck"){
  ret <- NULL
  for(file in files){
    cell_line <- gsub("^.*?Comp-(\\S+?)\\/.*$","\\1",file)
    if(!grepl("HCT116",cell_line))
      next
    if(type == "mageck"){
      td <- read.table(file, header=T, stringsAsFactors = F) %>%
        select(id, neg.score, neg.p.value, neg.fdr, neg.rank, neg.lfc, pos.score, pos.p.value, pos.fdr, pos.rank, pos.lfc) %>%
        rename(gene = id)
    }else{
      td <- read.delim(file, header=T, stringsAsFactors = T) %>%
        select(gene, TvCCvP_PositiveDesi, TvCCvP_NegativeDesi, TvC_negpval, TvC_pospval, TvC_LFC)
    }
    ret <- rbind(ret, data.frame(cell_line = cell_line, td), stringsAsFactors=F)
  }
  return(ret)
}


jaccard_index <- function(genes, data_1, data_2){
  jacc_num <- length(intersect(genes[data_1], genes[data_2]))
  jacc_denom <- length(union(genes[data_1], genes[data_2]))
  return(jacc_num/jacc_denom)
}


mh <- read_gene_summ(mageck_files)
dh <- read_gene_summ(desir_files, type = "desir")

```


# Hit counts

We do not see more candidate hits for the 3 replicate case.

```{r}
hit_comp <- mh %>%
  rowwise() %>%
  mutate(hit_fdr_5pct = (neg.fdr < 0.05 || pos.fdr < 0.05),
         hit_fdr_10pct = (neg.fdr < 0.1 || pos.fdr < 0.1)) %>%
  ungroup

hit_comp %>%
  group_by(cell_line) %>%
  summarise(Num_hits_fdr_5pct = sum(hit_fdr_5pct),
            Num_hits_fdr_10pct = sum(hit_fdr_10pct)) %>%
  ungroup %>%
  kable()

```

# Hit comparison

## Singleton similarities

The 3 replicate case shares more hits with the 2 replicate cases on the whole than they do with each other.

```{r}
ji <- pivot_wider(hit_comp, id_cols = c(gene,cell_line), 
                  names_from = cell_line, values_from = hit_fdr_10pct)
jd <- c(jaccard_index(ji$gene, ji$HCT116, ji$HCT116_2replicates_combn1),
        jaccard_index(ji$gene, ji$HCT116, ji$HCT116_2replicates_combn2),
        jaccard_index(ji$gene, ji$HCT116, ji$HCT116_2replicates_combn3),
        jaccard_index(ji$gene, ji$HCT116_2replicates_combn1, ji$HCT116_2replicates_combn2),
        jaccard_index(ji$gene, ji$HCT116_2replicates_combn1, ji$HCT116_2replicates_combn3),
        jaccard_index(ji$gene, ji$HCT116_2replicates_combn2, ji$HCT116_2replicates_combn3))
jacc.mat <- matrix(c(1,jd[1:3],
                         jd[1],1,jd[4:5],
                         jd[c(2,4)],1,jd[6],
                         jd[c(3,5,4)],1), 4, 4, byrow = T)
colnames(jacc.mat) <- rownames(jacc.mat) <- c("HCT116_3reps","HCT116_2reps-1","HCT116_2reps-2","HCT116_2reps-3")

ggcorr(jacc.mat, label = T, label_round = 3, cor_matrix = jacc.mat, limits=c(0.25,0.6),
       midpoint = 0.4) +
  labs(title = "HCT116 Hit similarities:",
       subtitle =  "Jaccard Index")

```

## Similarity to shared hits

The 3 replicate case does not exhibit a higher similarity to shared hits.

```{r}
js <- data.frame(Type = c("3_reps","2_reps.1","2_reps.2","2_reps.3"),
                 shared_hits.Jaccard_Index = c(jaccard_index(ji$gene, ji$HCT116, ji$gene %in% hits.shared$Gene),
                                               jaccard_index(ji$gene, ji$HCT116_2replicates_combn1, ji$gene %in% hits.shared$Gene),
                                               jaccard_index(ji$gene, ji$HCT116_2replicates_combn2, ji$gene %in% hits.shared$Gene),
                                               jaccard_index(ji$gene, ji$HCT116_2replicates_combn3, ji$gene %in% hits.shared$Gene)))

js %>%
  ggplot(aes(Type, shared_hits.Jaccard_Index, fill=Type)) +
  geom_bar(stat="identity") +
  ggtitle("HCT116: Similarity to shared hits")

```


# Power estimates: `locfdr`

Using Efron's local FDR approach (Efron 2007) and the accompanying R package `locfdr` to estimate the statistical power of 3 vs 2 replicates at an FDR of 10%. A summary statistic is used for each gene - below, we use p-values and log fold changes.

## P-values

```{r}
pv.3_reps <- mh$neg.p.value[mh$cell_line=="HCT116"]
pv.3_reps[pv.3_reps==1] <- 0.999
lf_3reps <- locfdr(qnorm(pv.3_reps), plot=0)
pv.2_reps.1 <- mh$neg.p.value[mh$cell_line=="HCT116_2replicates_combn1"]
pv.2_reps.1[pv.2_reps.1==1] <- 0.999
lf_2reps.1 <- locfdr(qnorm(pv.2_reps.1), plot=0)
pv.2_reps.2 <- mh$neg.p.value[mh$cell_line=="HCT116_2replicates_combn2"]
pv.2_reps.2[pv.2_reps.2==1] <- 0.999
lf_2reps.2 <- locfdr(qnorm(pv.2_reps.2), plot=0)
pv.2_reps.3 <- mh$neg.p.value[mh$cell_line=="HCT116_2replicates_combn3"]
pv.2_reps.3[pv.2_reps.3==1] <- 0.999
lf_2reps.3 <- locfdr(qnorm(pv.2_reps.3), plot=0)

roc <- rbind(as.data.frame(lf_3reps$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "3_reps"),
  as.data.frame(lf_2reps.1$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "2_reps.1"),
  as.data.frame(lf_2reps.2$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "2_reps.2"),
  as.data.frame(lf_2reps.3$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "2_reps.3"))


roc %>%
  ggplot(aes(FDR, Sensitivity, color = Type)) +
  geom_line() +
  ylim(0,1) +
  geom_vline(xintercept = 0.1, lintype = "dashed") +
  geom_abline(linetype = "dashed") +
  ggtitle("HCT116: replicate power comparison - p-values")

```

## Log-fold changes

```{r}
lf_3reps <- locfdr(scale(mh$neg.lfc[mh$cell_line=="HCT116"]), df = 15, plot=0)
lf_2reps.1 <- locfdr(scale(mh$neg.lfc[mh$cell_line=="HCT116_2replicates_combn1"]), df = 15, plot=0)
lf_2reps.2 <- locfdr(scale(mh$neg.lfc[mh$cell_line=="HCT116_2replicates_combn2"]), df = 15, plot=0)
lf_2reps.3 <- locfdr(scale(mh$neg.lfc[mh$cell_line=="HCT116_2replicates_combn3"]), df = 15, plot=0)

roc <- rbind(as.data.frame(lf_3reps$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "3_reps"),
  as.data.frame(lf_2reps.1$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "2_reps.1"),
  as.data.frame(lf_2reps.2$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "2_reps.2"),
  as.data.frame(lf_2reps.3$cdf1) %>%
  rename(FDR = p1, Sensitivity = V2) %>%
  mutate(Type = "2_reps.3"))


roc %>%
  ggplot(aes(FDR, Sensitivity, color = Type)) +
  geom_line() +
  ylim(0,1) +
  geom_vline(xintercept = 0.1, lintype = "dashed") +
  geom_abline(linetype = "dashed") +
  ggtitle("HCT116: replicate power comparison - logFC")

```

# References

Efron, B. 2007. Size, power and false discovery rates. Annals of Statistics 35: 1351-1377.

